---
title: "Predict the cost of homes in California"
author: "Clare Gibson"
date: "`r format(Sys.Date(), '%d %B %Y')`"
format:
  gfm:
    toc: true
    toc-depth: 2
---

# Introduction

This notebook trains a machine learning model that predicts home costs in California. The overall objective is to predict the value of home prices using **9 feature variables and 1 target variable**.

## Packages

The code chunk below loads all of the packages required for this exercise.

```{r}
#| label: load-packages
#| warnings: false
#| output: false

# Load packages
library(here)
library(tidyverse)
library(janitor)
```

## Data

The data for this exercise is contained in a csv file `housing.csv`.

```{r}
#| label: load-data
#| output: false

# Read data into a data frame
housing_df <- read_csv(here("data/src/housing.csv"))
```

# Exploratory analysis

## Glimpse

```{r}
#| label: glimpse-data

# Show column labels and data types
glimpse(housing_df)
```

## Head

```{r}
#| label: head-data

# Show the first few rows of the df
head(housing_df)
```

## Summary

```{r}
#| label: summary-data

# Show summary stats
summary(housing_df)
```

# Process data

## Missing data

First we take a look at which columns have missing data that could skew the model.

```{r}
#| label: view-missing-data

# Count NA values in each column
housing_df |> 
  summarise_all(~ sum(is.na(.))) |> 
  pivot_longer(everything())
```

The only column that contains missing values is `total_bedrooms`. It is not possible for a housing unit to have null bedrooms. We should be able to impute a value for the observations that are missing by using the k-nearest neighbours approach.
