---
title: "Linear Regression"
subtitle: "Applying linear regression methods using the diamonds dataset"
author: "Clare Gibson"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
# Knitr chunk options
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

# Introduction
In this notebook, I work through the steps of linear regression as taught in the [Machine Learning Specialization](https://www.coursera.org/specializations/machine-learning-introduction) available through [Coursera](https://coursera.org).

## Packages
I aim to write out most of the functions and calculations needed for linear regression using base or tidyverse-flavoured R, and not by using the specific machine learning packages that are available. However, I will make use of a number of R packages for data wrangling and plotting. These are all contained within the `tidyverse` wrapper. I also use a few other helpful packages as listed in the code below.

```{r load-packages}
# Load packages
library(tidyverse)
library(here)
```


## Data
For this exercise I use the diamonds dataset that is provided with 