---
title: "Linear Regression"
subtitle: "Applying linear regression methods using the diamonds dataset"
author: "Clare Gibson"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
# Knitr chunk options
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

# Introduction
In this notebook, I work through the steps of linear regression as taught in the [Machine Learning Specialization](https://www.coursera.org/specializations/machine-learning-introduction) available through [Coursera](https://coursera.org).

## Packages
I aim to write out most of the functions and calculations needed for linear regression using base or tidyverse-flavoured R, and not by using the specific machine learning packages that are available. However, I will make use of a number of R packages for data wrangling and plotting. These are all contained within the `tidyverse` wrapper. I also use a few other helpful packages as listed in the code below.

```{r load-packages}
# Load packages
library(tidyverse)
library(here)
```

## Data
For this exercise I use the diamonds dataset that is provided with the `ggplot2` package. This is a dataset containing the prices and other attributes of over 50,000 round cut diamonds. Since I have already loaded the `ggplot2` package (it is part of the `tidyverse` package), I can access the diamonds dataset by simply calling `diamonds`.

```{r show-diamonds-data}
diamonds
```

# Exploratory analysis
For this exercise, I will investigate whether I can predict the price of a diamond using one or more features. To start with, I will check if any of the observations have missing values. If so, they will need to be removed.

```{r count-missing-values}
sum(is.na(diamonds))
```

This dataset has no missing values so no need to remove any rows. Next, I will generate a series of plots to show the relationships between the `price` variable and other variables in the dataset[^1].

[^1]: Thanks to [this article](https://drsimonj.svbtle.com/plot-some-variables-against-many-others) for helping me to achieve this plot.

```{r plot-price}
# Plot the price variable against all other numerical variables
diamonds |> 
  pivot_longer(
    c(carat, depth, table, x, y, z),
    names_to = "var",
    values_to = "value"
  ) |> 
  ggplot(aes(x = value, y = price, color = cut)) +
    geom_point() +
    facet_wrap(~ var, scales = "free") +
    theme_bw()
```

From this plot, it appears that `price` has a positive linear relationship with `carat`, `x`, `y` and `z`.

## Train and test data

I will extract 80% of the dataset to use as training data and the remaining 20% will be the test data. In order to do this, I need to assign a unique ID to each row, then randomly select 80% of the rows using the `slice()` function. Using the row IDs, I can then find the remaining 20% to put into the test dataset.

```{r train-and-test-data}
# Add a unique ID to diamonds
df <- diamonds |> 
  rowid_to_column()

# Create training dataset
train <- df |> 
  slice_sample(prop = 0.8) |> 
  arrange(rowid)

# Create testing dataset
test <- df |> 
  filter(!rowid %in% train$rowid) |> 
  arrange(rowid)
```

# Univariate linear regression

Univariate linear regression uses a single feature $x$ to predict a value $y$. In this example, I will try to predict the price of a diamond from its carat value.

## Variables and parameters

- $x$ = carat
- $y$ = price
- $w,b$ = the parameters of the linear regression model

The code below loads the data into variables `x_train` and `y_train`.

```{r create-x-and-y-train}
x_train <- train$carat
head(x_train)

y_train <- train$price
head(y_train)
```

