---
title: "Linear Regression"
subtitle: "Applying linear regression methods using the diamonds dataset"
author: "Clare Gibson"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 10
    pandoc_args: --mathjax
---

```{r setup, include=FALSE}
# Knitr chunk options
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)
```

# Introduction
In this notebook, I work through the steps of linear regression as taught in the [Machine Learning Specialization](https://www.coursera.org/specializations/machine-learning-introduction) available through [Coursera](https://coursera.org).

## Packages
I aim to write out most of the functions and calculations needed for linear regression using base or tidyverse-flavoured R, and not by using the specific machine learning packages that are available. However, I will make use of a number of R packages for data wrangling and plotting. These are all contained within the `tidyverse` wrapper. I also use a few other helpful packages as listed in the code below.

```{r load-packages}
# Load packages
library(tidyverse)
library(skimr)   # for viewing summary statistics
library(here)    # for file location

# Load custom functions
source(here("R/utils.R"))
```

## Data
For this exercise I use the diamonds dataset that is provided with the `ggplot2` package. This is a dataset containing the prices and other attributes of over 50,000 round cut diamonds. Since I have already loaded the `ggplot2` package (it is part of the `tidyverse` package), I can access the diamonds dataset by simply calling `diamonds`.

```{r show-diamonds-data}
diamonds
```

# Exploratory analysis
For this exercise, I will investigate whether I can predict the price of a diamond using several features.

## Check for missing data
To start with, I will check if any of the observations have missing values. If so, they will need to be removed.

```{r count-missing-values}
sum(is.na(diamonds))
```

This dataset has no missing values so no need to remove any rows.

## Plot relationships
Next, I will generate a series of plots to show the relationships between the `price` variable and other variables in the dataset[^1].

[^1]: Thanks to [this article](https://drsimonj.svbtle.com/plot-some-variables-against-many-others) for helping me to achieve this plot.

```{r plot-price}
# Plot the price variable against all other numerical variables
diamonds |>
  pivot_longer(
    c(carat, depth, table, x, y, z),
    names_to = "var",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, y = price, color = cut)) +
    geom_point() +
    facet_wrap(~ var, scales = "free") +
    theme_bw()
```

From this plot, it appears that `price` has a positive linear relationship with `carat`, `x`, `y` and `z`.

## Set train and test data

I will extract 80% of the dataset to use as training data and the remaining 20% will be the test data. In order to do this, I need to assign a unique ID to each row, then randomly select 80% of the rows using the `slice()` function. Using the row IDs, I can then find the remaining 20% to put into the test dataset.

```{r train-and-test-data}
# Add a unique ID to diamonds
df <- diamonds |> 
  rowid_to_column()

# Create training dataset
train <- df |> 
  slice_sample(prop = 0.8) |> 
  arrange(rowid)

# Create testing dataset
test <- df |> 
  filter(!rowid %in% train$rowid) |> 
  arrange(rowid)
```

# Prepare data

Linear regression uses one or more features $\vec{x}$ to predict a value $y$. In this example, I will try to predict the price of a diamond from its features of `carat`, `x`, `y` and `z`.

## Steps for regression
The model function for linear regression is represented as:

$$
f_{\vec{w},b}(\vec{x})=\vec{w}\cdot\vec{x}+b
$$

where:

- $\vec{x}$ = [carat, x, y, z]
- $y$ = price
- $\vec{w},b$ = the parameters of the linear regression model

To train the linear regression model I want to find the best $\vec{w},b$ parameters that fit my dataset. I can evaluate how well a choice of $\vec{w},b$ fits the data by using the cost function $J(\vec{w},b)$.

To find the values $\vec{w},b$ that gets the smallest possible cost $J(\vec{w},b)$, I will use a method called gradient descent. 

## Create `x_train` and `y_train`
The code below loads the data into variables `x_train` and `y_train` and displays summary statistics about each.

```{r create-x-and-y-train-for-univariate}
x_train <- train |> 
  select(carat, x, y, z) |> 
  as.matrix()

head(x_train)
skim(x_train)

y_train <- train |> 
  select(price) |> 
  as.matrix()

head(y_train)
skim(y_train)
```

My dataset has `r nrow(x_train)` data points.

## Scale the `x_train` data
The features I am using for this regression exercise have different ranges.

To optimize my linear regression function I will scale the features so that they all fall within similar ranges.

```{r scale features}
# Zscore feature scaling
x_train_scaled <- scale(x_train)
```

# Regression

## Compute cost
In order to find the best values for $w$ and $b$ I need to try a lot of different values and calculate the cost of each attempt. Then using gradient descent I can obtain the optimal set of paramters to use in the linear regression function.

The cost function is defined mathematically as:

$$J(w,b)=\frac{1}{2m}\sum_{i=1}^{m}{(f_{w,b}(x^{(i)})-y^{(i)})^{2}}$$

```{r compute-cost}
# Try some starting variables
initial_w <- 0.2
initial_b <- -0.5

cost <- compute_cost(x_train_scaled, y_train, initial_w, initial_b)
print(class(cost))
paste0("Cost at initial w,b: ", cost)
```

## Compute gradient
The formulas to compute the gradients with respect to $w$ and $b$ are:
$$
\frac{\partial J(w,b)}{\partial b}  = \frac{1}{m} \sum\limits_{i = 0}^{m-1} (f_{w,b}(x^{(i)}) - y^{(i)})
$$

and

$$
\frac{\partial J(w,b)}{\partial w}  = \frac{1}{m} \sum\limits_{i = 0}^{m-1} (f_{w,b}(x^{(i)}) -y^{(i)})x^{(i)}
$$

## Gradient descent
In the next code cell I use the gradient descent function.

```{r gradient-descent}
# initialise parameters
initial_w <- matrix(0.0,,dim(x_train_scaled)[2])
initial_b <- 0.0
iterations <- 100
alpha <- 1.0e-1

# run gradient descent
# add this code to measure the time elapsed
start <- Sys.time()
grad_desc_list <- gradient_descent(
  x = x_train_scaled,
  y = y_train,
  w_in = initial_w,
  b_in = initial_b,
  cost_function = compute_cost,
  gradient_function = compute_gradient,
  alpha = alpha,
  num_iters = iterations
)
print(Sys.time() - start)
```

I can view how the gradient descent is performing by plotting the cost history by iteration.

```{r plot-gradient-descent}
# create a df containing iteration and cost history
j_hist <- tibble(
  iteration = grad_desc_list$iteration,
  cost = grad_desc_list$j_hist
)

# plot the cost values
j_hist |> 
  ggplot(aes(x = iteration, y = cost)) +
  geom_line()
```

# Prediction
The goal of my regression is to be able to predict the price of a diamond. I will use the linear regression function I just created to predict the price of a diamond that is 0.75 carat and has x, y, z values of 5.0, 5.7, 3.2.

First I need to normalize this test data then run it through tbe linear regression algorithm.